Effective Reionization Likelihood from Planck 2018 Data (?)

### 2021-03-12 paper/rei.tex

\section{Background}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}. We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.

%
%
 \begin{figure}
          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$. We assume full ionized Universe at $z < 6$. \textit{Bottom}: The cumulative optical depth resulting from a unit amplitude PC.} 
          \label{fig:xe}
\end{figure}

@@

 In Fig.~\ref{fig:xe}
 %
 \footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
 %
 (top), we show the corresponding fiducial basis functions $S_a(z)$.  
Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  
\wh{Possibly add a few words on the PC version and refer to the
bottom panel of Fig.~\ref{fig:xe}}
The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a  prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
% I think it can't depend on the data but does depend on the truncation
%the Planck 2018 data \wh{and
our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.}

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries and the unshaded region for $m_1$. The black solid line indicates the tanh models with varying $\tau$. The red and black crosses indicate the best-fit tanh models from Planck 2015 and 2018 respectively. Note that the best-fit tanh model has moved to a lower $\tau$ in Planck 2018, and in particular, is now within the 68\% C.L. contours of the constraints. 
\wh{double check relative sign on the definition of $m_i$ between 2015-18 - seems like some correlations have flipped, e.g. $m_4$.} }
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$
%
\footnote{We The Planck 2015 best-fit is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Plnack 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $dz = 0.015(1+z)$ instead of $dz = 0.5$ for the Planck 2015 best-fit. Using $dz = 0.5$ for Planck 2018 moves the cross negligible on this plot.}.
%
The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.  \wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}

@@

\begin{table}[b]
\centering
\caption{Planck 2015 results (with Mortonson PCs) [Will not be in the paper, just for us to compare] }
\label{tab:PC_stats}
\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
\hline

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
\caption{Example ionization history $x_e(z)$ for the high-$z$ model of \refsec{example2}. In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$. This is a toy model constructed for illustration purposes only. Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (XX, XX)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (XX, XX)$ (see also Fig.~\ref{two_parameter_model_2}). \ch{Include 2nd helium in the plot.}
}
\label{fig:two_step_model}
\end{figure}

@@

%\begin{figure}
%\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
%\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo. The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$. The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
}
%\label{fig:two_parameter_model_2D_plot}
%\end{figure}

@@

 We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
\beq
\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
\eeq
where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
The PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
the first two components
\beq
\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
\eeq
%
% Therefore, while retaining all physical models, they also includes some unphysical models. The authors of Ref.~\cite{Millea:2018bko} (hereafter MB18) raised a concern that these PC priors implies a non-flat prior in $\tau$ and claimed that it biases the PC result of  Ref.~\cite{Heinrich:2016ojb} and accounts for most of the shift between the tanh and PC $\tau$ posteriors. In Ref.~\cite{Heinrich:2018btc}, we showed that the use of our physicality prior does not cause bias for data that is constraining, which includes the Planck 2015 dataset. On the other hand, we showed in Ref.~\cite{Heinrich:2016ojb} that if one adopts the method proposed in Ref.~\cite{Millea:2018bko}. in an attempt to flatten the $\tau$ prior (by multiplying its inverse at each point in the multidimensional PC space), one would actually end up with a biased $\tau$ constraint for the Planck 2015 data. 
%
%We briefly recap the argument below. Let us use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily. We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
%\beq
%\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
%\eeq
%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model. Because most of the contributions to $\tau$ comes from the first two components (see also Fig. 7 of Ref.~\cite{Heinrich:2018btc} in the Appendix), we can approximate it as
%\beq
%\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
%\eeq
Due to this linearity, flat priors in $m_1$ and $m_2$ correspond to
flat priors in $\tau_{12}$ for any linear trajectory through the space.   For example, the tanh model trajectory is shown in Fig.~{\ref{fig:plot_mjs_2018_vs_2015}.   Even without the explicit 
construction of a flat tanh $\tau$ prior in \S \ref{sec:example1}, 
the implicit prior from $m_1,m_2$ is already nearly flat. \ch{CH: need to ask a question about this.}

Ref.~\cite{Millea:2018bko} raised a potential objection regarding more general models.  As illustrated in Fig.~\ref{fig:prior_box}, lines of constant $\tau_{12}$ are rotated compared with the $m_1$,
$m_2$ axis.  Therefore, the range-bound flat prior in $m_1,m_2$ (blue square) considered {\it alone} corresponds to a prior that is not flat in $\tau_{12}$ due to the extent  of the allowed parameter space
for each of its values, e.g. at $\tau_{12}=0.018$
than at $\tau_{12} = 0.125$.   Ref.~\cite{Millea:2018bko} suggest 
inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in the \ch{Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}}.
First notice that this inversion would in fact then produce a non-flat prior on $\tau$ for the tanh model
that disfavors high values.   This is because
the additional parameter space allowed by the range-bound $m_a$ prior is excluded by the functional form of the model even though the reweighting occurs globally and includes the tanh trajectory as well.

@@

The real impact of our range-bound $m_1,m_2$ priors comes from the fact that with the Planck 2018 data, the physicality prior on $m_1$
removes nearly half of the space that would otherwise be allowed by the
data at 95\% CL.  This is in contrast to the Planck 2015 data where only a very small region is removed (see Fig.~\ref{fig:plot_m1m2_2015_vs_2018}).  \wh{Maybe add to the $\tau$ vs $\tau_{12}$ figure a version that does not have this $\tau_{12}$ cut.  Discuss more if the difference significant}.

As discussed in \S \ref{sec:cumulative}, our range-bound priors 
are conservative in the sense that all the models that are excluded are definitely unphysical but some of the models that are included
may also be unphysical since their physicality depends on higher order PC components that are explicitly truncated in the analysis.  On the other hand the tanh model provides an explicitly physical example whose trajectory runs nearly parallel to the our physicality prior and so we do not expect the bias incurred by 
including these models to be large. \ch{CH: sorry, what do we mean by these models here?}  In any case, using PCs to 
test explicitly physical models as in \S \ref{sec:example1} and
\ref{sec:example2} cannot be affected by our conservative physicality
cut.

@@

 \begin{figure}
          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau_prior_fac_0.8.pdf}
          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) \ch{CH: [why is this flat in m1 and m2?]} which are aligned with lines of constant $\tau_{12}$ (light gray lines). Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses). The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.} 
          \label{fig:prior_box}
\end{figure}

@@

\item $\tau = 0.0487+0.0038
−0.0081$ (lowE; flat $\tau$ prior; PCA)(68\%).

>>>>

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}. We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.

%
%
 \begin{figure}
          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$. We assume fully ionized Universe at $z < 6$. \textit{Bottom}: The cumulative optical depth resulting from a unit amplitude PC. \ch{plot done}} 
          \label{fig:xe}
\end{figure}

@@

 In Fig.~\ref{fig:xe}
 %
 \footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
 %
 (top), we show the corresponding fiducial basis functions $S_a(z)$.  
Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from contraints on the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  
\wh{Possibly add a few words on the PC version, list $\tau_a = \tau_a(z_{\rm min},z_{\rm max})$ in an equation or table, and refer to the
bottom panel of Fig.~\ref{fig:xe}}
The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a  prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
% I think it can't depend on the data but does depend on the truncation
%the Planck 2018 data \wh{and
our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries and the unshaded region for $m_1$. The black solid line indicates the tanh models with varying $\tau$. The red and black crosses indicate the best-fit tanh models from Planck 2015 and 2018 respectively. Note that the best-fit tanh model has moved to a lower $\tau$ in Planck 2018, and in particular, is now within the 68\% C.L. contours of the constraints. \ch{plot done}
}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$
%
\footnote{We The Planck 2015 best-fit is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Plnack 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $dz = 0.015(1+z)$ instead of $dz = 0.5$ for the Planck 2015 best-fit. Using $dz = 0.5$ for Planck 2018 moves the cross negligible on this plot.}.
%
The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.  \wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}

@@

\begin{table}[b]
\centering
\caption{$\tau_a$}
\label{tab:PC_stats}
%\begin{tabular}{|r | r r r r r r|}
\begin{tabular}{|c | c | c | c | c | c|}
\hline
$\tau_{\rm fid}$ & $\tau_1$  & $\tau_2$ & $\tau_3$ & $\tau_4$ & $\tau_5$ \\ \hline
0.08626 & 0.29403 & $-$0.11227 & 0.04506 & $-$0.01898 & 0.00960 \\
\hline
\end{tabular}
\end{table}
% all digits without flipping m3, m4:
%tau_fid 0.08625526676478058
%taumj [ 0.29403315 -0.11227135 -0.04505736  0.0189787   0.00960184
\begin{table}[b]
\centering
\caption{Planck 2015 results (with Mortonson PCs) [Will not be in the paper, just for us to compare] }
\label{tab:PC_stats}
\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
\hline

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
\caption{Example ionization history $x_e(z)$ for the high-$z$ model of \refsec{example2}. In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$. This is a toy model constructed for illustration purposes only. Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (XX, XX)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (XX, XX)$ (see also Fig.~\ref{two_parameter_model_2}  \wh{place x's on the models plotted in that plot}). \ch{Include 2nd helium in the plot.}
}
\label{fig:two_step_model}
\end{figure}

@@

%\begin{figure}
%\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
%\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo. The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$. The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
%}
%\label{fig:two_parameter_model_2D_plot}
%\end{figure}

@@

 We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
\beq
\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
\eeq
where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
The PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
the first two components
\beq
\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
\eeq
\wh{possibly redefine
\beq
\tau_{12} =  m_1 \tau_1 + m_2 \tau_2 +
\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) \approx \tau.
\eeq
where the term in parentheses is a fixed quantity (refer to eqns/table where tau's and mbars are defined) whose purpose is to remove offsets due to the arbitrary choice of fiducial model.
}
%
% Therefore, while retaining all physical models, they also includes some unphysical models. The authors of Ref.~\cite{Millea:2018bko} (hereafter MB18) raised a concern that these PC priors implies a non-flat prior in $\tau$ and claimed that it biases the PC result of  Ref.~\cite{Heinrich:2016ojb} and accounts for most of the shift between the tanh and PC $\tau$ posteriors. In Ref.~\cite{Heinrich:2018btc}, we showed that the use of our physicality prior does not cause bias for data that is constraining, which includes the Planck 2015 dataset. On the other hand, we showed in Ref.~\cite{Heinrich:2016ojb} that if one adopts the method proposed in Ref.~\cite{Millea:2018bko}. in an attempt to flatten the $\tau$ prior (by multiplying its inverse at each point in the multidimensional PC space), one would actually end up with a biased $\tau$ constraint for the Planck 2015 data. 
%
%We briefly recap the argument below. Let us use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily. We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
%\beq
%\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
%\eeq
%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model. Because most of the contributions to $\tau$ comes from the first two components (see also Fig. 7 of Ref.~\cite{Heinrich:2018btc} in the Appendix), we can approximate it as
%\beq
%\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
%\eeq
Due to this linearity, flat priors in $m_1$ and $m_2$ correspond to
flat priors in $\tau_{12}$ for any linear trajectory through the space.   For example, the tanh model trajectory is shown in Fig.~{\ref{fig:plot_mjs_2018_vs_2015}.   Even without the explicit 
construction of a flat tanh $\tau$ prior in \S \ref{sec:example1}, 
the implicit prior from $m_1,m_2$ is already nearly flat. \ch{CH: need to ask a question about this.}

Ref.~\cite{Millea:2018bko} raised a potential objection regarding more general models.  As illustrated in Fig.~\ref{fig:prior_box}, lines of constant $\tau_{12}$ are rotated compared with the $m_1$,
$m_2$ axis.  Therefore, the range-bound flat prior in $m_1,m_2$ (blue square) considered {\it alone} corresponds to a prior that is not flat in $\tau_{12}$ due to the extent  of the allowed parameter space
for each of its values, e.g. at  left corners (Fig.~\ref{fig:prior_box} solid square or lowest $m_1$) of the range $\tau_{12}=0.018$
than at $\tau_{12} = 0.125$ \wh{revise numbers} and the implied prior increases linearly from zero between the two.
   Ref.~\cite{Millea:2018bko} suggest 
inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in the \ch{Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}}.
First notice that this inversion would in fact then produce a non-flat prior on $\tau$ for the tanh model
that disfavors high values.   This is because
the additional parameter space allowed by the range-bound $m_a$ prior is excluded by the functional form of the model even though the reweighting occurs globally and includes the tanh trajectory as well.

@@

The real impact of our range-bound $m_1,m_2$ priors comes from the fact that with the Planck 2018 data, the physicality prior on $m_1$
removes nearly half of the space that would otherwise be allowed by the
data at 95\% CL.  This is in contrast to the Planck 2015 data where only a very small region is removed (see Fig.~\ref{fig:plot_m1m2_2015_vs_2018}).  \wh{Maybe add to the $\tau$ vs $\tau_{12}$ figure a version that does not have this $\tau_{12}$ cut.  Discuss more if the difference significant - also mention that even in the direct tanh constraints, the impact of taking $z\gtrsim 6$ has the effect of removing some of the low $\tau$ region.}.

As discussed in \S \ref{sec:cumulative}, our range-bound priors 
are conservative in the sense that all the cases that are excluded are definitely unphysical but some of the cases that are included
may also be unphysical since their physicality depends on higher order PC components that are explicitly truncated in the analysis.  On the other hand the tanh model provides an explicitly physical example whose trajectory runs close  and nearly parallel to the our physicality prior in the allowed region and so we do not expect the bias incurred by 
including these possibly \wh{unphysical cases} to be large. 
%\ch{CH: sorry, what do we mean by these models here?}  
In any case, using PCs to 
test explicitly physical models as in \S \ref{sec:example1} and
\ref{sec:example2} cannot be affected by our conservative physicality
cut.

@@

 \begin{figure}
          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau_prior_fac_0.8.pdf}
          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
           $\tau_{12}$ and \wh{range-bound flat - also if we change $\tau_{12}$ definitions change the numbers in the labels here} in $m_1-m_2$ (rotated rectangle, dashed lines) 
           %\ch{CH: [why is this flat in m1 and m2?]} 
           % flat in 2D or locally flat
           which are aligned with lines of constant $\tau_{12}$ (light gray lines). Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses). The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.} 
          \label{fig:prior_box}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
\caption{$\tau_{12,new}$}
}
\label{fig:tau12}
\end{figure}

@@

\item $\tau = 0.0487+0.0038
-0.0081$ (lowE; flat $\tau$ prior; PCA)(68\%).
